### 2023-09-03 sections/1-introduction.tex
merge conflict

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters; 
the GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences.

More recently, Big-ANN's NeurIPS'23: Billion Scale Approximate Nearest Neighbor Search Challenge has 
solicits development of data structures and search algorithms for Approximate Nearest Neighbor (ANN) search
on four massive datasets: YFCC 100M~\cite{johnson2019billion} dataset, the Yandex Text-to-Image 10M~\cite{diskann-github}, cross-modal dataset, 
the MSMARCO passage retrieval dataset~\cite{Bruch2023AnAA}, and a 10M slice of the MS Turing data set~\cite{Singh2021FreshDiskANNAF}. These datasets 
are 192, 200, 30,000, and 100 dimensional respectively, and submissions are evaluated with 
a ten million point sample of each.

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating 
that computer architecture will not "catch up" to our computational needs in the near future. As a result, analysis of these large datasets 
requires better algorithms.

Many researchers are especially interested in performing similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, similarity search can become a useful tool for 
understanding and verifying the validity of model outputs. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus speed tradeoff~\cite{ishaq2019clustered}.

@@

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{indyk1999sublinear}, 
clever indexing techniques such as the FM Index~\cite{simpson2010efficient}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the geometric and topological structure inherent in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CAKES (CLAM-Accelerated $K$-nearest-neighbor 
Entropy-scaling Search). CLAM (Clustered Learning of Approximate Manifolds), developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the CHESS clustering algorithm. 
Using the cluster tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform
$k$-nearest neighbors search.

This paper focuses on four novel exact $k$-nearest neighbors search algorithms. We also explain some 
implementation improvements to the clustering and $\rho$ nearest neighbors search algorithms in CHESS, 
as well as improved genericity across distance functions. Additionally, we prove a guarantee about the scaling behavior of cluster radii. 
Finally, we provide a comparison of 
CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion} on several datasets 
from the ANN and Big ANN benchmarks suites.

>>>>

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters; 
the GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences. [tTODO: add dimensionality here]

% More recently, Big-ANN's NeurIPS'23: Billion Scale Approximate Nearest Neighbor Search Challenge has 
% solicits development of data structures and search algorithms for Approximate Nearest Neighbor (ANN) search
% on four massive datasets: YFCC 100M~\cite{johnson2019billion} dataset, the Yandex Text-to-Image 10M~\cite{diskann-github}, cross-modal dataset, 
% the MSMARCO passage retrieval dataset~\cite{Bruch2023AnAA}, and a 10M slice of the MS Turing data set~\cite{Singh2021FreshDiskANNAF}. These datasets 
% are 192, 200, 30,000, and 100 dimensional respectively, and submissions are evaluated with 
% a ten million point sample of each.

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating 
that computer architecture will not ``catch up'' to our computational needs in the near future. As a result, analysis of these large datasets 
requires better algorithms.

Many researchers are especially interested in similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, GPT-4, and LLAMA-2, [CITE] similarity search can become a useful tool for 
understanding and verifying the validity of model embeddings. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus throughput tradeoff~\cite{ishaq2019clustered}.

@@

[TODO: talk about need for exact search in a medical setting; talk to Noah about this, good strains of bacteria, drug response curves]

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{indyk1999sublinear}, 
clever indexing techniques such as the FM Index~\cite{simpson2010efficient}, hierarchical navigable small world (HNSW) networks [CITE], and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the geometric and topological structure inherent in large datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CAKES (CLAM-Accelerated $K$-nearest-neighbor 
Entropy-scaling Search). CLAM (Clustered Learning of Approximate Manifolds), developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the CHESS clustering algorithm. 
Using the cluster tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform
$k$-nearest neighbors search.

This paper focuses on four novel algorithms for exact $k$-NN search. We also present some 
improvements to the clustering and $\rho$-nearest neighbors search algorithms in CHESS, 
as well as improved genericity across distance functions. Additionally, we prove a guarantee about the scaling behavior of cluster radii. 
Finally, we provide a comparison of 
CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion} and HNSW, on several datasets 
from the ANN benchmarks suite [CITE].

### 2023-09-03 sections/2-methods.tex
typo
merge conflict

In this manuscript, we are primarily concerned with $k$-nearest neighbors in a finite-dimensional 
vector space. Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the vector representation of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
When, in addition to these constraints, the distance function obeys
the triangle inequality, it is a \emph{distance metric}, and in such cases we can guarantee that our search algorithms have perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point 
and length scale in the dataset, 
which we define as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold at a given point, for some length scale.
The notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space is known as the \emph{manifold hypothesis}~\cite{fefferman2016testing}.

Figures 2 and 3 in~\cite{ishaq2019clustered} illustrate the low fractal dimensions of two datasets used with $\rho$-NN search in CHESS. 
We observe that for both APOGEE (Figure 2 in~\cite{ishaq2019clustered}) and GreenGenes (Figure 3 in~\cite{ishaq2019clustered}), other than the most extreme 10\% of clusters, virtually all clusters have a local fractal dimension of less than $2 \ll 8,575$
and $2 \ll 2,250$ respectively.
This suggests that APOGEE and GreenGenes are good datasets for use with entropy-scaling search algorithms like CHESS and CAKES.

@@

CLAM assumes the manifold hypothesis. 
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy 
a $d$-dimensional manifold, where $d \ll D$. 
While we sometimes use Euclidean notions, such as voids and volumes, to talk about geometric and topological 
properties of clusters and of the manifold, CLAM does not rely on such notions; 
they serve merely as a convenient, more intuitive way to talk about the underlying mathematics.

\subsubsection {Cluster Partitioning}

For a cluster $C$ with $|C|$ points, we begin by computing a 
random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances 
between all points in the sample. Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
that is, the point which minimizes the sum of distances to all other points in the sample. This geometric median 
is $C$'s \emph{center}. The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}, and the point $r$ which is furthest
from $l$ is designated the \emph{right pole}. We then partition the cluster into a \emph{left child} and \emph{right child}, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or some other user-specified stopping criterion 
is met. This process is described in Algorithm \ref{alg:partition}.

@@

\subsubsection {Dataset Depth-First Reordering}
\label{subsubsec:methods:dataset-depth-first-reordering}

CAKES also improves upon CHESS by reordering the dataset. In CHESS, each cluster stored a list of indices into the dataset, 
which we used during search to retrieve a cluster's points. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices for each of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
With $\log{}n$ levels in the tree, this approach had $\mathcal{O}(n\log{}n)$ memory cost.

One may think to improve this memory cost to $\mathcal{O}(n)$ by storing only the indices of points at 
the leaf cluster level.
This approach, however, introduces an $\mathcal{O}(n)$ time cost whenever we need to find the indices for 
a non-leaf cluster, since it requires a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset 
so that points are stored in a depth-first order. Then, within each cluster, we need only 
store the cluster's cardinality and an \emph{offset} to access the cluster's points from the dataset. Each parent cluster 
have the same offset as its left child, and each right child has an offset equal to their sibling's offset
plus their siblings's cardinality. With no additional memory cost nor time cost for retrieving points during search, 
dataset depth-first reordering offers significant improvements over the approach used in CHESS.

\subsubsection {Scaling Behavior of Cluster Radii}
\label{subsubsec:methods:guaranteed-decrease-in-cluster-radii}
While it may be tempting to assume that cluster radii decrease with each application of Cluster Partition (refer to Algorithm \ref{alg:partition}), unfortunately, this assumption is incorrect. 
Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii; in particular, we prove in this section that cluster radii will stop increasing after at most 
$d$ partitions, where $d$ is the dimensionality of the dataset.

We can describe a $d$-dimensional distribution of data by choosing some set of $d$ mutually orthogonal axes.
Let $2R$ be the maximum pairwise distance among the points in the dataset. 
We choose the axes such that the two points that are $2R$ apart lie \emph{along one of the axes}. 
Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset. 
In the worst case, (i.e., with a uniform-density distribution that fills the $d$-sphere), our axes will be such that $2R$ is the maximum pairwise distance along every axis. 
Such a distribution would also produce a balanced clustering.

In this case, Cluster Partition will select a maximally distant pair of points to use as poles, i.e., it will choose one of the $d$ axes along
which to split the cluster into two children. 
After one application of Algorithm \ref{alg:partition}, the maximum pairwise distance along that axis will be
bounded above by $R$. 
The next recursive Cluster Partition will select another of the $d$ axes. 
Thus, after at most $d$ applications of Algorithm \ref{alg:partition}, the
maximum pairwise distance along each axis will be bounded above by $R$. 
The overall (i.e., not restricted to one axis) maximum pairwise distance 
will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
See Figure 1 for a concrete example.

Thus, starting with a cluster $C$ of radius $R$, after at most $d$ Partitions, the descendants of $C$ will each have radius
bounded above by $\frac{R}{\sqrt{2}}$. In other words, cluster radii are guaranteed to decrease by a multiplicative factor of $\frac{\sqrt{2}}{2}$ after at 
most $d$ applications of Algorithm \ref{alg:partition}.

Note that, in practice, we never see a balanced clustering. Cluster Partition  produces unbalanced trees due to the varying density of the sampling 
in different regions of the manifold and the low dimensional "shape" of the manifold. Further, the cluster radii decrease by a factor much larger than 
$\frac{\sqrt{2}}{2}$ in practice, and the upper bound of $d$ partitions is seldom realized.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
    \caption{Scaling behavior of radii on a uniform-density disk, which in two dimensions represents the worst-case scenario. 
    $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
    After one application of Algorithm~\ref{1}, we have $C_1$, with radius $R_1$ and center $o_1$.
    The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 \leq R_1$.
    Hence, the radius of a child cluster can be larger than its parent. However, 
    after another application of Algorithm~\ref{1}, we have consumed all $d$ orthogonal axes, 
    as shown in $C_2$. 
    Now, clearly $R_2 \leq R_0$. In fact, for each subsequent application of Algorithm~\ref{1},
    the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.}
    \label{fig:methods:scaling_behavior}
\end{figure}

@@

The asymptotic complexity of Cluster Partition is the same as described in~\cite{ishaq2019clustered}
By using an approximate partitioning with a $\sqrt{n}$ sample, we achieve $\mathcal{O}(n)$ cost of 
partitioning and $\mathcal{O}(n \log n)$ cost of building the tree. This is a significant improvement over 
exact partitioning and tree-building, which have $\mathcal{O}(n^2)$ and $\mathcal{O}(n^2 \log n)$ costs respectively.

\subsection {Sharding and Auto-Tuning}
\label{subsec:methods:index-building}
In this work, we also introduce the method of \emph{sharding} for datasets that cannot fit in memory. 
This approach also leverages the fact that $\rho$-nearest neighbors search is typically much faster than $k$-nearest neighbors search.
In addition to sharding, we perform some naive auto-tuning to select the optimal $k$-NN algorithm to be used with a given dataset.

With this approach, we begin by taking a random sample of up to 10\% of the data. 
The exact proportion of the dataset in this random sample depends on the dimensionality of the dataset and a user-specified memory limit. 
FIrst, we build a cluster tree on this sample, and 
then consider all clusters at some user-specified depth in the tree. Using the cluster centers 
of these clusters as queries, and a user-specified value of $k$, we perform $k$-NN search using each of the four algorithms described in Section~\ref{subsec:methods:knn-search}. 
We record the 
time taken for each algorithm and select the one which had the lowest total search time over all the queries. 
We then split the rest of the dataset into some user-specified number of shards, and use the algorithm identified in the previous step to perform $k$-NN search on the first shard. 
This search finds the $k$ closest neighbors to the query \emph{in the first shard}. 
We take the furthest of those $k$ neighbors and use its distance from the query as the radius to be used with $\rho$-nearest neighbors search (see Section~\ref{subsec:methods:rnn-search}) on each remaining shard. 
We then take the union of the results of these $\rho$-nearest neighbors searches, along with the $k$ neighbors from the first shard, and then linear search over those points to find the actual $k$ nearest neighbors.

Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we use the selected algorithm to perform search on the given dataset for any value of $k$.
Note also that the size of the shards is not necessarily consistent; it is often optimal to have the first shard be smaller than the others, since the first shard is the only one for which we perform $k$-nearest neighbors search. 
However, making this first shard too small introduces the risk of significantly overestimating the radius (i.e., the distance to the $k$-th ), resulting in much slower $\rho$-nearest neighbors search on the subsequent shards.
Finding the precise optimal ratio between the size of the first shard and the size of the remaining shards, as well as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the actual $k$-th closest neighbor, are topics for future work.

For datasets which \emph{do} fit in memory, we use the same approach to auto-tuning, but with only one shard.

@@

Putting our bounds for coarse and fine search together, we have that $k$-NN by Repeated $\rho$-NN has 
$\mathcal{O}\left(\log_2{\mathcal{N}_{\hat{r_c}}(D)} + k\left(1+ \frac{2\hat{r_c}}{\hat{r}}\left(\frac{k}{\hat{C}}\right)^{\frac{-1}{\hat{d}}}\right)^d\right) = 
\mathcal{O}\left(\log_2{\mathcal{N}_{\hat{r_c}}(D)} + k\left(\frac{2\hat{r_c}}{\hat{r}}\right)^d\left(\frac{k}{\hat{C}}\right)^{\frac{-d}{\hat{d}}}\right)$ 
time complexity, where $\mathcal{N}_{\hat{r_c}}(D)$ is the metric entropy of the dataset, $d$ is the dimensionality of the dataset, and $k$ is the number of nearest neighbors.

>>>>

In this manuscript, we are primarily concerned with $k$-nearest neighbors in a finite-dimensional space. Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the neural embedding of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \cup \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number. We require that the distance function 
be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
In addition to these constraints, if the distance function obeys
the triangle inequality, i.e., it is a \emph{distance metric}, then we guarantee that CAKES has perfect recall. 
Choice of distance function varies by type of data. For example, with electromagnetic spectra, we use both 
Euclidean (L2) and Cosine distances. With biological or string data, Hamming and Levenshtein distances are more appropriate.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point 
and length scale in the dataset. 
We define local fractal dimension (LFD) as: 
\begin{equation} \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) } \label{1} \end{equation}
where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ 
centered at a point $q$ in the dataset $\textbf{X}$; in this example, we compute fractal dimension for some radius $r_1$ and a smaller radius $r_2 = \frac{r_1}{2}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset. To illustrate the difference,
consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575. Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional 
manifold within the 8,575-dimensional embedding space. The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
This notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space is known as the \emph{manifold hypothesis}~\cite{fefferman2016testing}.

Figures 2 and 3 in~\cite{ishaq2019clustered} illustrate the low fractal dimensions of two datasets used with $\rho$-NN search in CHESS. 
We observe that for both APOGEE (Figure 2 in~\cite{ishaq2019clustered}) and GreenGenes (Figure 3 in~\cite{ishaq2019clustered}), other than the most extreme 10\% of clusters, virtually all clusters have a local fractal dimension of less than 2. 
This suggests that APOGEE and GreenGenes are good datasets for use with entropy-scaling search algorithms like CHESS and CAKES.

@@

CAKES assumes the manifold hypothesis. 
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy 
a $d$-dimensional manifold, where $d \ll D$. 
While we sometimes use Euclidean notions, such as voids and volumes, to talk about geometric and topological 
properties of clusters and of the manifold, CLAM does not rely on such notions; 
they serve merely as a convenient, intuitive way to talk about the underlying mathematics.

[TODO: change to just partition because I took about quicksort partition]
\subsubsection {Cluster Partitioning}

For a cluster $C$ with $|C|$ points, we begin by computing a 
random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances 
between all points in the sample. Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
that is, the point which minimizes the sum of distances to all other points in the sample. This geometric median 
is $C$'s \emph{center}. The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}. The point $r$ which is furthest
from $l$ is designated the \emph{right pole}. We then partition the cluster into a \emph{left child} and \emph{right child}, where the 
left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all 
points in the cluster which are closer to $r$ than to $l$. Without loss of generality, we assign to the left child 
those points which are equidistant from $l$ and $r$. Starting from a root-cluster containing the entire dataset, we 
repeat this procedure until each leaf contains only one datum or some other user-specified stopping criterion 
is met. This process is described in Algorithm \ref{alg:partition}.

@@

\subsubsection {Depth-First Reordering}
\label{subsubsec:methods:dataset-depth-first-reordering}

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.
In CHESS, each cluster stored a list of indices into the dataset, 
which we used during search to retrieve a cluster's points. 
Though this approach allowed us to retrieve the points in constant 
time, its memory cost was prohibitively high; since each cluster stores indices for each of its
points, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
Assuming $\log{}n$ levels in the tree, this approach had $\mathcal{O}(n\log{}n)$ memory cost. 
[TODO: talk to Noah about how to phrase this, how to get at that balanced tree is the worst case]

One may improve this memory cost to $\mathcal{O}(n)$ by storing only these indices at 
the leaf cluster level.
This approach, however, introduces an $\mathcal{O}(n)$ time cost whenever we need to find the indices for 
a non-leaf cluster, since it requires a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset 
so that points are stored in a depth-first order. Then, within each cluster, we need only 
store the cluster's cardinality and an \emph{offset} to access the cluster's points from the dataset. A left child has 
the same offset as its parent cluster, and the corresponding right child has an offset equal to the left child's offset
plus the left child's cardinality. With no additional memory cost nor time cost for retrieving points during search, 
dataset depth-first reordering offers significant improvements over the approach used in CHESS.

\subsubsection {Scaling Behavior of Cluster Radii}
\label{subsubsec:methods:guaranteed-decrease-in-cluster-radii}
While it may be tempting to assume that cluster radii decrease with each application of Partition (refer to Algorithm \ref{alg:partition}), unfortunately, this assumption is incorrect. 
Fortunately, we \emph{can} make some guarantees about the scaling behavior of cluster radii; in particular, we prove in this section that cluster radii will stop increasing after at most 
$d$ partitions, where $d$ is the fractal dimensionality of the dataset.

We can describe a $d$-dimensional distribution of data by choosing some set of $d$ mutually orthogonal axes.
Let $2R$ be the maximum pairwise distance among the points in the dataset. 
We choose the axes such that the two points that are $2R$ apart lie \emph{along one of the axes}. 
Thus, a $d$-dimensional hyper-sphere of radius $R$ would bound the dataset. 
In the worst case, (i.e., with a uniform-density distribution that fills the $d$-sphere), our axes will be such that $2R$ is the maximum pairwise distance \emph{along every axis}. 
Such a distribution would also produce a balanced cluster tree.

In this case, Partition will select a maximally distant pair of points to use as poles; without loss of generality, 
say it chooses two points at the extrema one of the $d$ axes. 
After one partition, the maximum pairwise distance along that axis will be
bounded above by $R$. 
The next recursive Partition will select another of the $d$ axes, and once again, the distance along that axis will be bounded above by $R$
Thus, after at most $d$ partitions, the
maximum pairwise distance along each axis will be bounded above by $R$. 
The overall (i.e., not restricted to one axis) maximum pairwise distance 
will be bounded above by $R\sqrt{2}$ by, for example, two instances that lie at the extrema of different axes. 
See Figure~\ref{fig:methods:scaling_behavior} for an example.

Thus, starting with a cluster $C$ of radius $R$, after at most $d$ Partitions, the descendants of $C$ will each have radius
bounded above by $\frac{R}{\sqrt{2}}$. In other words, cluster radii are guaranteed to decrease by a multiplicative factor of $\frac{\sqrt{2}}{2}$ after at 
most $d$ partitions.

Note that, in practice, we never see a balanced clustering. Cluster Partition  produces unbalanced trees due to the varying density of the sampling 
in different regions of the manifold and the low dimensional ``shape'' of the manifold. Further, in practice, the cluster radii decrease by a factor much larger than 
$\frac{\sqrt{2}}{2}$ with every partition; the upper bound of $d$ partitions is seldom realized.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/geometry/geometry.pdf}
    \caption{Scaling behavior of radii on a uniform-density disk, which represents the worst-case scenario for a two-dimensional distribution. 
    $R_0$ denotes the radius of the root cluster $C_0$, and $o_0$ denotes its center. 
    After one application of Algorithm~\ref{1}, we have $C_1$, with radius $R_1$ and center $o_1$.
    The right triangle formed by $o_0$, $o_1$, and $y_+$ in $C_1$ shows that $R_0 \leq R_1$.
    Hence, the radius of a child cluster can be larger than its parent. However, 
    after another application of Algorithm~\ref{1}, we have consumed all $d$ orthogonal axes, 
    as shown in $C_2$. 
    Now, clearly $R_2 \leq R_0$. In fact, for each subsequent application of Algorithm~\ref{1},
    the radius of the resulting cluster is bounded above by $\frac{\sqrt{2}R}{2}$.}
    \label{fig:methods:scaling_behavior}
\end{figure}

@@

The asymptotic complexity of Partition is the same as described in~\cite{ishaq2019clustered}
By using an approximate partitioning with a $\sqrt{n}$ sample, we achieve $\mathcal{O}(n)$ cost of 
partitioning and $\mathcal{O}(n \log n)$ cost of building the tree. This is a significant improvement over 
exact partitioning and tree-building, which have $\mathcal{O}(n^2)$ and $\mathcal{O}(n^2 \log n)$ costs respectively.

\subsection {Sharding and Auto-Tuning}
\label{subsec:methods:index-building}
In this work, we also introduce a method of \emph{sharding} for datasets that cannot fit in memory. 
This approach also leverages the fact that $\rho$-nearest neighbors search at the radius of the $k$-th farthest neighbor 
is typically much faster than $k$-nearest neighbors search.
In addition to sharding, we perform some naive auto-tuning to select the optimal $k$-NN algorithm to be used with a given dataset.

With this approach, we begin by taking a random sample of no more than 10\% of the data. This sample 
is considered the first shard. 
The exact proportion of the dataset in this random sample depends on the dimensionality of the dataset and a user-specified memory limit. 
First, we build a cluster tree on this sample, and 
then consider all clusters at some user-specified depth in the tree, with a default depth of 10. Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search using each of the four algorithms described in Section~\ref{subsec:methods:knn-search}. We select the algorithm which had the lowest total search time over all the queries. 
We then split the rest of the dataset into shards based on the user-defined memory limit.

Sharded search then proceeds as follows. We use previously determined $k$ algorithm on the first shard. 
We use the distance from the $k$-th farthest neighbor to perform $\rho$-NN search on each subsequent shard. 
After searching each shard, we keep only the $k$-nearest neighbors so far, potentially decreasing the distance 
from the $k$-th farthest neighbor for search on the next shard.

Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.
Note also that the size of the shards is not necessarily consistent; it is often optimal to have the first shard be smaller than the others, since the first shard is the only one for which we perform $k$-nearest neighbors search. 
However, making this first shard too small introduces the risk of significantly overestimating the radius (i.e., the distance to the $k$-th neighbor), resulting in much slower $\rho$-nearest neighbors search on the subsequent shards.
Finding the precise optimal ratio between the size of the first shard and the size of the remaining shards, as well as computing the expected level of error between the distance from the $k$-th closest neighbor in the first shard and the actual $k$-th closest neighbor, are topics for future work. For datasets which \emph{do} fit in memory, we use the same approach to auto-tuning, but, since the entire dataset fits in memory, there is only one shard.

@@

Putting our bounds for coarse and fine search together, we have that $k$-NN by Repeated $\rho$-NN has 
$\mathcal{O}\left(\log_2{\mathcal{N}_{\hat{r_c}}(D)} + k\left(1+ \frac{2\hat{r_c}}{\hat{r}}\left(\frac{k}{\hat{C}}\right)^{\frac{-1}{\hat{d}}}\right)^d\right) = 
\mathcal{O}\left(\log_2{\mathcal{N}_{\hat{r_c}}(D)} + k\left(\frac{2\hat{r_c}}{\hat{r}}\right)^d\left(\frac{k}{\hat{C}}\right)^{\frac{-d}{\hat{d}}}\right)$ 
time complexity, where $\mathcal{N}_{\hat{r_c}}(D)$ is the metric entropy of the dataset, $d$ is the dimensionality of the dataset, and $k$ is the number of nearest neighbors. We note that the fraction $\frac{-d}{\hat{d}}$ should, unless fractal dimension is highly variable (that is, $\hat{d}$ differs significantly from $d$), be relatively close to 1.
